Measuring causal influence with\\ back-to-back regression

### 2019-05-22 paper/conclusion.tex

In this work, we explored the back-to-back (B2B) regression, a novel method for
identifying causes from observations building on two successive
multidimensional regressions; from one domain to the other, and then back.
We analyzed the asymptotic behavior of this approach, proving its convergence,
and compared it to several baselines in controlled synthetic experiments. We
showcased our approach on a recent brain imaging dataset, analyzing the timing
of brain responses and their connection to features of words, and obtained
results consistent with prior work in neuroscience literature, confirming the
viability of our method in the case of real data.

>>>>

In this work, we proposed Back-to-Back Regression (B2BR), a method to 
measure the causal influence of a potential set of variables generating some observations.
%
B2BR performs two successive multidimensional regressions: one from the output domain, and another one from the input domain.
%
We provided a theoretical guarantee about the consistency of B2BR, and compared it to several baselines in controlled synthetic experiments.
%
We also applied B2BR to a recent brain imaging dataset, analyzing the timing
of brain responses and their connection to word features.
%
We obtained results consistent with prior work in neuroscience literature, confirming the efficacy of B2BR for real data analysis.

### 2019-05-22 paper/introduction.tex

First, causal and non-causal factors may be numerous and collinear. In physics, for example, a student may have difficulty identifying whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience for example, it can be challenging to identify whether the frequency of a word presented on a subject' retina modulates brain activity. Indeed, word frequency strongly covaries with word length (short words are more frequent than long words) as well as with word category (determinants are more frequent than adverbs), the individual letters from which it is composed etc (e.g. words that contain "z" are rare) and so forth. As it can be particularly difficult to generate a set of words that control for all of these factors simultaneously, it is common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, etc), and investigate the resulting parameters of the best model.

The second challenge to causal discovery is that observations are becoming increasingly large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus.

As illustrated in Fig. 1., both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors. \par

In this paper, we introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling. After detailing the method and proving its convergence, we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of causal discovery and out-of-environment predictions. Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to remarkably distinguishable brain responses.

>>>>

First, causal and non-causal factors may be numerous and collinear. In physics, for example, a student may have difficulty identifying whether fusion is caused by a change in temperature or in pressure, as these two factors can be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity. Indeed, word frequency strongly covaries with word length (short words are more frequent than long words) as well as with word category (determinants are more frequent than adverbs), the individual letters from which it is composed etc (e.g. words that contain "z" are rare) and so forth. As it can be particularly difficult to generate a set of words that control for all of these factors simultaneously, it is common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, etc), and investigate the resulting parameters of the best model.

The second challenge to measuring causal influence is that observations are becoming increasingly large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use 'decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus.
%
Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors. \par

In this paper, we introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling. After detailing the method and proving its convergence, we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating causal influence and out-of-environment predictions. Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to remarkably distinguishable brain responses.

### 2019-05-22 paper/paper.tex

The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (FK)^\dagger$.

The second step is straightforward, replacing Y, we have
\begin{equation}
\begin{aligned}
\arg \min_H \left \| \hat GY - HX \right \|^2 &=\arg  \min_H \left \| \hat GF(EX+N) - HX \right \|^2 \\
&=\arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2 + \left \| \hat GFN \right \| ^2\\
&= \arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2\\
&=\hat G (FE)
\end{aligned}
\end{equation}

Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} S (FK)^{\dagger}(FKK') = \frac{1}{1+nsr} E$.

As we shall see, the diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E. Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
% \subsection{Asymptotic behaviour - non linear case}

>>>>

The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (KF)^\dagger$.

The second step is straightforward, replacing Y, we have
\begin{equation}
\begin{aligned}
\arg \min_H \left \| \hat YG - XH \right \|^2 &=\arg  \min_H \left \| (XE + N)F \hat G - XH \right \|^2 \\
&=\arg \min_H \left \| X(EF \hat G - H) \right \| ^2 + \left \| NF\hat G \right \| ^2\\
&= \arg \min_H \left \| X(EF \hat G - H) \right \| ^2\\
&= EF \hat G 
\end{aligned}
\end{equation}

Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.

%The diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E. Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
% \subsection{Asymptotic behaviour - non linear case}

### 2019-05-22 paper/synthetic.tex

Table~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods...

\clearpage
\newpage

>>>>

Table~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods.
%
B2BR is the method obtaining best test error out-distribution, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.
